A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-12-24 paper/paper.tex
paper: minor polish

A \emph{presheaf} on an category is a family of sets indexed by the object of the category with maps indexed by the morphisms. As such, it lives on the indexed side of the equivalence, contrasting with the fibered side, where we have \emph{discrete Grothendieck fibrations}~\citep{LoregianRiehl20}. However, there are situations where a presheaf $F$ can also be seen as living on the fibered side of the equivalence. This happens when the indexing category is \emph{direct}, or has a downwards-well-founded collection of non-identity morphisms.Let us consider, for instance, the case of a semi-cubical set~\citep{grandis03,buchholtz17} presented with $2n$ face maps from the set of $n$-cubes to the set of $(n-1)$-cubes. Formulated in type theory, the corresponding presheaf definition of a semi-cubical set prescribes a family of sets and face maps between them as follows.

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
up to cubical faces identities. Here, $X_1$ can be seen as a family over $X_0 \times X_0$ presented in a fibered form, $X_2$ can be seen as a family over $X_1 \times X_1 \times X_1 \times X_1$. This suggests an alternative indexed presentation of the presheaf as a stratified sequence of families indexed by families of lower rank, taking into account the coherence conditions to prevent duplications. Formulated in type theory, it takes the form:
\begin{align*}
  X_0 & :              & \U                                                                            \\
  X_1 & :              & X_0 \times X_0 \rightarrow \U                                                 \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \times X_1 (c)(d) \times X_1(a)(c) \times X_1 (b)(d) \rightarrow \U \\
  \ldots
\end{align*}

The idea for such an indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types\footnote{See e.g. \href{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}{ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A few constructions have been proposed since then. The first construction by \cite{voevodsky12} relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. The second, by \cite{herbelin15}\footnote{In hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somewhat confusing: it claimed to construct semi-simplicial types, but the construction was done in a type theory with Unicity of Identity Proofs. Consequentially, what was really obtained was an indexed presentation of semi-simplicial sets. The confusion was however, common at the time.} formalized in the Coq proof assistant, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by \cite{part15} formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17} formalized in (an emulation of a two-level type theory in) the Agda proof assistant\footnote{\href{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}{github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

@@

Our work is a step in the direction of the program initiated in \cite{altenkirch15} to develop parametricity-based models of parametric type theory~\citep{bernardy15,nuyts17,cavallo19} and cubical type theory~\citep{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory, and is likely to better reflect the definitional properties of type theory than presheaf-based cubical sets would. For example, consider the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~\citep{curien14}.

@@

More generally, the standard augmented $(n + 1)$-semi-simplex can be obtained by taking a copy of the standard augmented $n$-semi-simplex serving as a base and gluing on top of it another copy lifted by one dimension. In the second copy, the color becomes an extra point, the points become lines connecting the points of the base to the extra point, and so on. In particular, the components of the base are those of the standard augmented $n$-semi-simplex postfixed by $0$ while the components of the lifted copy are postfixed by $\kstar$. Note that the components may be oriented by letting each $n$-dimensional component point to the $(n-1)$-dimensional component obtained by replacing the leftmost $\kstar$ of the $n$-dimensional component with $0$.

@@

\section{Future work}
The construction could be extended with degeneracies as well as with permutations~\citep{grandis03}. Dependent $\nu$-sets could also be defined, opening the way to construct $\Pi$-types and $\Sigma$-types of $\nu$-sets. A $\nu$-set of $\nu$-sets representing a universe could also be defined (as sketched e.g. in a talk at the HoTT-UF workshop for the bridge case~(\citeyear{herbelin-hott-uf})). More generally, we believe these lines of work would eventually provide alternative models to parametric type theories~\citep{nuyts17,cavallo19} where equality of types, now a family rather than the total space of a fibration, is not only definitionally isomorphic to bridges~\citep{bernardy15}, but definitionally the same as bridges.

>>>>

A \emph{presheaf} on an category is a family of sets indexed by the object of the category with maps indexed by the morphisms. As such, it lives on the indexed side of the equivalence, contrasting with the fibered side, where we have \emph{discrete Grothendieck fibrations}~\citep{LoregianRiehl20}. However, there are situations where a presheaf $F$ can also be seen as living on the fibered side of the equivalence. This happens when the indexing category is \emph{direct}, or has a downwards-well-founded collection of non-identity morphisms. Let us consider, for instance, the case of a semi-cubical set~\citep{grandis03,buchholtz17} presented with $2n$ face maps from the set of $n$-cubes to the set of $(n-1)$-cubes. Formulated in type theory, the corresponding presheaf definition of a semi-cubical set prescribes a family of sets and face maps between them as follows.

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
up to cubical faces identities. Here, $X_1$ can be seen as a family over $X_0 \times X_0$, and $X_2$ can be seen as a family over $X_1 \times X_1 \times X_1 \times X_1$, in the fibered presentation. This suggests an alternative indexed presentation of the presheaf as a stratified sequence of families indexed by families of lower rank, taking into account the coherence conditions to prevent duplications. Formulated in type theory, it takes the form:
\begin{align*}
  X_0 & :              & \U                                                                            \\
  X_1 & :              & X_0 \times X_0 \rightarrow \U                                                 \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \times X_1 (c)(d) \times X_1(a)(c) \times X_1 (b)(d) \rightarrow \U \\
  \ldots
\end{align*}

The idea for such an indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types\footnote{\href{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}{ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}}. A few constructions have been proposed since then. The first construction by \cite{voevodsky12} relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. The second, by \cite{herbelin15}\footnote{In hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somewhat confusing: it claimed to construct semi-simplicial types, but the construction was done in a type theory with Unicity of Identity Proofs. Consequentially, what was really obtained was an indexed presentation of semi-simplicial sets. The confusion was however, common at the time.} formalized in the Coq proof assistant, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by \cite{part15} formalized in an emulation of logic-enriched homotopy type theory in the Plastic proof assistant, and yet another by \cite{annenkovCK17} formalized in an emulation of a two-level type theory in the Agda proof assistant\footnote{\href{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}{github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

@@

Our work is a step in the direction of the program initiated in \cite{altenkirch15} to develop parametricity-based models of parametric type theory~\citep{bernardy15,nuyts17,cavallo19} and cubical type theory~\citep{bezem13,cohen16,angiuli21}, which are closer to the syntax of type theory, and are likely to better reflect the definitional properties of type theory than presheaf-based cubical sets would. For example, consider the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~\citep{curien14}.

@@

More generally, the standard augmented $(n + 1)$-semi-simplex can be obtained by taking a copy of the standard augmented $n$-semi-simplex serving as a base, and gluing on top of it another copy lifted by one dimension. In the second copy, the color becomes an extra point, the points become lines connecting the points of the base to the extra point, and so on. In particular, the components of the base are those of the standard augmented $n$-semi-simplex postfixed by $0$ while the components of the lifted copy are postfixed by $\kstar$. Note that the components may be oriented by letting each $n$-dimensional component point to the $(n-1)$-dimensional component obtained by replacing the leftmost $\kstar$ of the $n$-dimensional component with $0$.

@@

\section{Future work}
The construction could be extended with degeneracies as well as with permutations~\citep{grandis03}. Dependent $\nu$-sets could also be defined, opening the way to construct $\Pi$-types and $\Sigma$-types of $\nu$-sets. A $\nu$-set of $\nu$-sets representing a universe could also be defined as sketched in a talk at the HoTT-UF workshop for the bridge case~(\citeyear{herbelin-hott-uf}). More generally, we believe these lines of work would eventually provide alternative models to parametric type theories~\citep{nuyts17,cavallo19} where equality of types, now a family rather than the total space of a fibration, is not only definitionally isomorphic to bridges~\citep{bernardy15}, but definitionally the same as bridges.
